---
title: "Edinburgh llord reg"
author: "nb"
date: "29 March 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install packages - first time only
# packs <- c("tidyverse", "dplyr", "readxl")
# install.packages(packs)

# load libraries
library(dplyr)
library(tidyverse)
library(readxl)




```

# Landlord register - City of Edinburgh 


## FOI dataset
Data were obtained from City of Edinburgh Council under a Freedom of Information (FoI) request (GIVE DETAILS). These data provide the number of listed properties within each unit postcode in the city at xxx. All the postcodes begin with "EH" as expected. 

One has 9 characters rather than the expect 7 or 8 but this appears to be a due to an additional space as is easily corrected. However, it is surprising that the system does not seem to check and prevent invalid postcodes from being entered. 

```{r get FOI data}
# open csv file from FoI request
df_foi <- as_tibble(read_excel("data/16177 Response Data.xlsx"))

# rename cols to remove spaces
names(df_foi) <- c("PropertyPostcode", "NumberOfProperties")

# make id column from row number
df_foi <- tibble::rowid_to_column(df_foi, "ID")

# make pc8 as uppercase version of original - name to match var in pcode lookup
df_foi <- mutate(df_foi, 
                 pc8 = toupper(PropertyPostcode))
       
# this shows that first two letters in pc8 are always EH
table(str_sub(df_foi$pc8, 1, 2))

# this shows that length of pc8 mostly 7 or 8, but one 9
table(str_length(df_foi$pc8))

# identify the case - "EH16 4 FE"
subset(df_foi, str_length(pc8) > 8)

# change that case
df_foi$pc8[df_foi$pc8 == "EH16 4 FE"] <- "EH16 4FE"

# check that it is fixed
table(str_length(df_foi$pc8))

# quick check on final state of df_foi
head(df_foi)

```

In total, there are `r nrow(df_foi)` postcodes with registered properties and a total of `r sum(df_foi$NumberOfProperties)`. That represents an average of `r sum(df_foi$NumberOfProperties)/nrows(df_foi)' properties per postcode. 



## Postcode lookup file
We read in the postcode to higher area lookup file, cut this down to "EH" postcodes and to the 11 variables we are interested in. 


```{r get pcode lookup}

# open postcode directory - gives some errors which we ignore for now
df_pcode <- read_csv(
  "data/Scottish_Postcode_Directory_2017_2.csv",
  col_types = cols(
      pc7 = col_character(),
      pc8 = col_character(),
      Grid_Reference_Easting = col_double(),
      Grid_Reference_Northing = col_double(),
      Latitude = col_double(),
      Longitude = col_double(),
      DataZone2011 = col_character(),
      HHC2011 = col_double(),
      Pop2011 = col_double(),
      HHC2001 = col_double(),
      Pop2001 = col_double()
  )
)
# head(df_pcode)


# one cut down of pcodes to those starting EH and only the vars we need
df_pcode2 <- subset(df_pcode, str_sub(pc7, 1, 2) == "EH",
                   select = c(pc7, pc8, DataZone2011, LAU_Level1_2011
                   )
)

# second cut down of pcodes to just those in Edinburgh
df_pcode3 <- subset(df_pcode, LAU_Level1_2011 == "S30000008",
                   select = c(DataZone2011
                   )
)

# collapse so it is just the list of DZs in Edinburgh
df_dz_Edin <- df_pcode3 %>%
  group_by(DataZone2011) %>%
  summarise()
# extract as list
dz_Edin <- (df_dz_Edin$DataZone2011)

```

There are `r nrow(df_dz_Edin)` Datazones in Edinburgh in 2011. 


## Aggregate FoI data to Datazone
We use the postcode lookup file to attach 2011 Datazone identifiers to the FOI data, along with information on the population of the postcode sector. 


```{r aggregate to DZ}

# left-join data from df_pcode to df_foi
df_foi <- df_foi %>%
  left_join(df_pcode2, by = "pc8")
head(df_foi)

# check how many cases have missing values - for pc7 it is 18 
sum(is.na(df_foi$pc7))

# check number of rows and valid cases in all cols
# there is some drop off in HHC and Pop figs, esp. for 2001
nrow(df_foi)
colSums(!is.na(df_foi))

```

Number of cases: `r nrow(df_foi)`
Pcode lookup failed: `r sum(is.na(df_foi$pc7))`


# 



```{r FOI data by DZ}

# aggregating data to DZ level, summing number of registered props
df_foi_dz <- df_foi %>%
  group_by(DataZone2011) %>%
  summarise(
    NumberOfProperties = sum(NumberOfProperties, na.rm = TRUE)
    )

```


## Census data on Datazones


```{r census data for DZs}

# Open file with tenure composition for DZs in 2011
df_census <- as_tibble(read_csv("data/tenure DZ2011.csv"))

# make correct PRS and PRSPct figures
df_census <- mutate(df_census, 
                    PRS = PrtRent_2011, 
                    PRSPct = PrtRent_2011/All_HH_2011)

# reduce to DZs in Edinburgh (dz_Edin)
df_census <- df_census[df_census$DataZone %in% dz_Edin,]

# rename 'Datazone' as 'Datazone2011'
colnames(df_census)[1] <- "DataZone2011"
```


Now we merge tenure and landlord register data on to the Edinburgh Datazones file. 

```{r xx}

df_dz_Edin <- df_dz_Edin %>%
  left_join(df_census, by = "DataZone2011")
  
df_dz_Edin <- df_dz_Edin %>%
  left_join(df_foi_dz, by = "DataZone2011")

# check number of DZ and number with registered properties
nrow(df_dz_Edin)
sum(is.na(df_dz_Edin$NumberOfProperties))

# where reg props missing, set to zero
df_dz_Edin$NumberOfProperties[
  is.na(df_dz_Edin$NumberOfProperties)
  ] <- 0

```




```{r graphs}
# make text for correlation
x <- signif(
  cor(df_dz_Edin$PRS, df_dz_Edin$NumberOfProperties), 
  digits = 2)
corr_text <- paste("R =", x)

ggplot(data = df_dz_Edin, 
       mapping = aes(x = PRS, y = NumberOfProperties)) +
  geom_point() +
  geom_smooth(method = 'lm') + 
  ggtitle("Registered properties vs PRS households - Edinburgh") +
  xlab("PRS households in 2011 (Census)") +
  ylab("No. of registered properties in 201x") +
  geom_text(x = 25, y = 450,
            label = corr_text)



```



```{r xx1}

```



```{r xx2}

```


