---
title: "Data note: Edinburgh landlord register data analysis"
author: "Mark Livingston and Nick Bailey"
date: "April 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install packages - first time only
# packs <- c("tidyverse", "dplyr", "readxl")
# install.packages(packs)

# load libraries
library(dplyr)
library(tidyverse)
library(readxl)

```

# Introduction

Data were obtained from City of Edinburgh Council's register of private landlords. Data were obtained under a Freedom of Information (FoI) request in December 2017. (The request was made on 07/09/2017 - request number 16177). These data provide the number of listed properties within each unit postcode in the city at that time, i.e. they are an aggregated, anonymised extract from the database. In this note, we explore how these data can be used to provide a picture of private renting in the city.


# Other data on the PRS in Edinburgh

At the 2011 Census, there where 224,240 households in Edinburgh, of which 49,980 (22.3%) were in the private rented sector(PRS). We can estimate changes since then using the Scottish Survey Core Questions (SSCQ) for 2012-17. There has been a slight rise in the population of Edinburgh. In 2016, there were an estimated 232,900 households - an increase of 3.9% from 2011. The number of households in private renting rose to a peak of nearly 70,000 in 2014 but fell in subsequent years to 63,920 in 2016 (27.4%). 

[Mark - can we update to 2017? number and % PRS in 2012 (assuming 2011 not available) and number/% in 2017]


#FOI dataset
```{r get FOI data, include=FALSE}
# open csv file from FoI request
df_foi <- as_tibble(read_excel("data/16177 Response Data.xlsx"))

# rename cols to remove spaces
names(df_foi) <- c("PropertyPostcode", "NumberOfProperties")

# make id column from row number
df_foi <- tibble::rowid_to_column(df_foi, "ID")

# make pc8 as uppercase version of original - name to match var in pcode lookup
df_foi <- mutate(df_foi, 
                 pc8 = toupper(PropertyPostcode))
       
# this shows that first two letters in pc8 are always EH
table(str_sub(df_foi$pc8, 1, 2))

# this shows that length of pc8 mostly 7 or 8, but one 9
table(str_length(df_foi$pc8))

# identify the case - "EH16 4 FE"
subset(df_foi, str_length(pc8) > 8)

# change that case
df_foi$pc8[df_foi$pc8 == "EH16 4 FE"] <- "EH16 4FE"

# check that it is fixed
table(str_length(df_foi$pc8))

# quick check on final state of df_foi
head(df_foi)

```

Data were provided as a .csv file, listing each unit postcode containing any registered properties and the number of registered properties. It should be noted that, where a property is owned by two or more individuals, each is required to register separately. In this dataset, duplicate listings for the same property have been combined. 

Various checks were carried out on the postcodes. All begin with "EH" as expected. All but one has the expected 7 or 8 characters but one had 9. This is due to an additional space so is easily corrected. However, it is surprising that the system does not seem to check and prevent an obviously invalid postcode from being entered. 

In total, there are `r nrow(df_foi)` postcodes with registered properties and they contain a total of `r sum(df_foi$NumberOfProperties)`. That represents an average of `r sum(df_foi$NumberOfProperties)/nrows(df_foi)` properties per postcode. 

The number of registered properties is `r 100 * sum(df_foi$NumberOfProperties)/63920`% of the estimated total for Edinburgh (using the SSCQ figure from 2016). At best, there appear to be `r 100 - (100 * sum(df_foi$NumberOfProperties)/63920)`% of properties not listed (around `r 63920 - sum(df_foi$NumberOfProperties)` properties). That figure is almost certainly an underestimate since properties can be listed by not let. 


# Comparison with Census data at Datazone level

The great advantage of the landlord register data is that it provides detail down to the small area level. Here we aggregate data from unit postcodes to Datazones, and then use these to compare with Census data for the same units. 

```{r get pcode lookup, include=FALSE}

# open postcode directory - gives some errors which we ignore for now
df_pcode <- read_csv(
  "data/Scottish_Postcode_Directory_2017_2.csv",
  col_types = cols(
      pc7 = col_character(),
      pc8 = col_character(),
      Grid_Reference_Easting = col_double(),
      Grid_Reference_Northing = col_double(),
      Latitude = col_double(),
      Longitude = col_double(),
      DataZone2011 = col_character(),
      HHC2011 = col_double(),
      Pop2011 = col_double(),
      HHC2001 = col_double(),
      Pop2001 = col_double()
  )
)
# head(df_pcode)

# one cut down of pcodes to those starting EH and only the vars we need
df_pcode2 <- subset(df_pcode, str_sub(pc7, 1, 2) == "EH",
                   select = c(pc7, pc8, DataZone2011, LAU_Level1_2011
                   )
)

# second cut down of pcodes to just those in Edinburgh
df_pcode3 <- subset(df_pcode, LAU_Level1_2011 == "S30000008",
                   select = c(DataZone2011
                   )
)

# collapse so it is just the list of DZs in Edinburgh
df_dz_Edin <- df_pcode3 %>%
  group_by(DataZone2011) %>%
  summarise()
# extract as list
dz_Edin <- (df_dz_Edin$DataZone2011)

# left-join data from df_pcode to df_foi
df_foi <- df_foi %>%
  left_join(df_pcode2, by = "pc8")
head(df_foi)

# check how many cases have missing values - for pc7 it is 18 
sum(is.na(df_foi$pc7))

# check number of rows and valid cases in all cols
# there is some drop off in HHC and Pop figs, esp. for 2001
nrow(df_foi)
colSums(!is.na(df_foi))

# aggregating data to DZ level, summing number of registered props
df_foi_dz <- df_foi %>%
  group_by(DataZone2011) %>%
  summarise(
    NumberOfProperties = sum(NumberOfProperties, na.rm = TRUE)
    )

```
To aggregate unit postcodes to Datazones, we use a lookup file. In this process, the lookup failed in `r sum(is.na(df_foi$pc7))` cases - a very small fraction. 

There are `r nrow(df_dz_Edin)` Datazones in Edinburgh in 2011 and `r nrow(df_foi_dz)` of these have registered landlords in 2017. 


## Census data on Datazones


```{r census data for DZs, include=FALSE}

# Open file with tenure composition for DZs in 2011
df_census <- as_tibble(read_csv("data/tenure DZ2011.csv"))

# make correct PRS and PRSPct figures
df_census <- mutate(df_census, 
                    PRS = PrtRent_2011, 
                    PRSPct = PrtRent_2011/All_HH_2011)

# reduce to DZs in Edinburgh (dz_Edin)
df_census <- df_census[df_census$DataZone %in% dz_Edin,]

# rename 'Datazone' as 'Datazone2011'
colnames(df_census)[1] <- "DataZone2011"
```


Now we merge tenure and landlord register data on to the Edinburgh Datazones file. 

```{r xx, include=FALSE}

df_dz_Edin <- df_dz_Edin %>%
  left_join(df_census, by = "DataZone2011")
  
df_dz_Edin <- df_dz_Edin %>%
  left_join(df_foi_dz, by = "DataZone2011")

# check number of DZ and number with registered properties
nrow(df_dz_Edin)
sum(is.na(df_dz_Edin$NumberOfProperties))

# where reg props missing, set to zero
df_dz_Edin$NumberOfProperties[
  is.na(df_dz_Edin$NumberOfProperties)
  ] <- 0

```


A scatterplot shows the relationship between the number of properties in the PRS according to the Census 2011 and the number of registered properties in 2017. 

```{r graphs}
# make text for correlation
x <- signif(
  cor(df_dz_Edin$PRS, df_dz_Edin$NumberOfProperties), 
  digits = 2)
corr_text <- paste("R =", x)

ggplot(data = df_dz_Edin, 
       mapping = aes(x = PRS, y = NumberOfProperties)) +
  geom_point() +
  geom_smooth(method = 'lm') + 
  ggtitle("Registered properties vs PRS households - Edinburgh") +
  xlab("PRS households in 2011 (Census)") +
  ylab("No. of registered properties in 201x") +
  geom_abline(slope=1)
  geom_text(x = 25, y = 450,
            label = corr_text)

```
